Towards a standardized evaluation of imputation methodology

### 2022-11-17 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
added to text and table

One potential reason for discrepancies between imputation method performance in simulations versus applications is the use of sampling variance in the simulation design. Simulation studies on missing data methodology have the unique option to exclude conventional sampling schemes and focus on a single finite population. When a finite population is made incomplete, sampling variance must be excluded from inference evaluations and the missing data generation procedure serves as the necessary source of Monte Carlo variation \citep{vink14, raghunathan2003multiple}. \citet{vink14} propose pooling rules for multiple imputation on finite populations and demonstrate that valid inferences are then obtained by excluding the sampling variance components from the calculation of total variance about $(Q - \bar{Q})$, where $Q$ would be the estimand and $\bar{Q} = \sum_{l=1}^m \hat{Q}/m$ is the average over the $m$ imputed estimates \citep[cf.][p. 76-77]{rubi87}. The total variance $T=B+B/m$ would then solely rely on the between imputation variance $B = \sum_{l=1}^m (\hat{Q}_l - \bar{Q})^\prime(\hat{Q}_l - \bar{Q}) / m-1$.

@@

The challenge with model-based data-generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data hardly ever follow a known theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness would subsequently be induced, the observed parts of the data will still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good.

@@

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to mimic a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the true data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is generated randomly. Figure \ref{Fig} demonstrates this: Both univariate MAR mechanisms in Figure \ref{Fig} are induced based on the right tail, but only the mechanisms for which the \emph{variable to ampute} and the \emph{missingness covariate} have a positive correlation will yield valid MAR mechanisms. In the case of zero correlation between the data columns, iterative univariate non-MCAR missingness schemes may induce invalid missingness mechanisms that mimic MCAR. Table \ref{Tab} demonstrates this and highlights the importance of including complete case analysis as an evaluation analysis to allow for identifying such invalidly generated missingness mechanisms in simulation. In fact, this would amount to one of the special cases described by \citet{buur18} under which complete case analysis would be more efficient than imputation: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

@@

% latex table generated in R 3.2.2 by xtable 1.8-0 package
% Thu Feb  4 16:27:54 2016
\begin{table}[t!]
\caption{Inferences obtained by complete case analysis (CCA) and multiple imputation over 1000 simulations. Displayed are the bias of the mean, coverage rate (cov) of the corresponding 95\% confidence interval and the confidence interval width (ciw). Imputations are generated by Bayesian linear regression imputation.}
\centering
\begin{tabular}{rlccccccc}
  \hline
 	& &\multicolumn{3}{c}{CCA} &&\multicolumn{3}{c}{imputation}	\\\cline{3-5}\cline{7-9}
 \multicolumn{2}{c}{mechanism}	& bias & cov & ciw && bias & cov & ciw \\ 
  \hline
  $[\rho = .8]$ &mcar & 0.001 & 0.959 & 0.179 && -0.001 & 0.946 & 0.156 \\ 
  $[\rho = .8]$ &mar & {\color{red} 0.334} & {\color{red} 0.000} & 0.169 && 0.004 & 0.955 & 0.195 \\ 
  $[\rho = .0]$ &mar & 0.003 & 0.973 &{\color{red} 0.180} && 0.013 & 0.947 & 0.311 \\    \hline
\end{tabular}
\label{Tab}
\end{table}

A non-ignorable or MNAR mechanism--where the unobserved values themselves may also play a role in the probability to be missing--might fall outside many studies' scope, yet could yield informative insights for daily practice. Even if an imputation method is specifically developed with ignorable missingness in mind, chances are that the method will be applied to non-ignorable empirical missingness at some point in time. After all, for every MNAR mechanism there is a MAR mechanism with equal fit and one cannot definitively verify that empirical missingness is random \citep{molenberghs2008every}. Therefore, it can be argued that MNAR is the more likely mechanism for real-life missingness scenarios. It may be wise to include MNAR missingness in the simulation study just in case: A method that performs well under MAR and some cases of MNAR may be preferable over a method that only works under MAR, and not under MAR. Such `late-phase` methodological investigations \citep[see][]{hei22} can provide an understanding of an imputation method's robustness. Alternatively, a sensitivity analysis may provide an indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid \citep[see e.g.][part 5]{molenberghs2014handbook}.

@@

At minimum, imputation method performance should be quantified using appropriate measures. It depends on the specifics of each simulation study which performance measures would be most suitable. If the goal is inference--not prediction--the uncertainty about estimates needs to be properly quantified. To capture this uncertainty, the standard errors of the estimates should be correctly calculated, which requires multiple imputation. The aim of multiple imputation is not to reproduce the data, but to allow for obtaining valid inference given that the data are incomplete. This means that, given the framework provided by \citet{rubi87}, statistical properties such as bias, confidence intervals, and the coverage rate of the confidence intervals should always be studied. We therefore recommend evaluating the following points:

(i) The methods should preferably be unbiased. Note that the way bias is calculated should be carefully chosen and described, since this can greatly influence the interpretation of the results. A negligible absolute bias, for example, may already yield an almost infinite relative bias if the true value of the estimand is zero. In most cases, unbiased estimates may be expected under a MCAR missingness mechanism, which can be easily verified by including MCAR missingness in the scope of missingness investigations.

(ii) The intervals for estimates should have a valid coverage of the population (i.e. true) value. Coverage of a 95\% interval should in theory be $\geq 95$\%, where a coverage rate of 95\% would be most efficient \citep[p. 589]{neym34}. Under-coverage indicates that the procedure is not confidence valid and may lead to invalid inference. Under-coverage may occur when the estimation procedure is either biased, too liberal leading to intervals that are too narrow, or both. Over-coverage, though technically not confidence invalid, would indicate that efficiency could still be gained. In such cases, a narrower interval exists for which the method would still yield confidence valid result.

@@

(iv) Predictive accuracy may be quantified using the root mean squared error (RMSE). We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values and it may inflate the type I error rate of statistical inferences \citep[Chapter 2.6][]{buur18}. However, if a study is aimed at obtaining predictions, and inferential validity is not of interest, the RMSE of the predicted values can be compared between methods. Then, the RMSE may yield valuable information about the methods' comparative efficiency in terms of accuracy and precision.

After establishing the statistical validity of an imputation routine on the level of the analysis estimates, it remains paramount to also study the imputation-generating process. Omitting such investigations may yield sub-optimal results. One simple example arises often in practice: generating implausible imputed values. Even though the estimates on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny. Under many circumstances, imputation methods may be realistically expected to preserve both marginal and conditional distributions with respect to the comparative truth. Imputation methods that fail to do so, should not be considered general-purpose methods. Many techniques can yield valid inferences, but techniques that sample realistic or plausible values may be preferable in practice.

@@

\begin{table}[ht!]
\caption{Checklist for reporting on imputation methodology evaluations.}
\label{table:check}
\begin{tabular}{ll}
\hline
$\square$ & \textbf{Aims}                                                                                  \\
  & e.g. simulation scope                                                                          \\ \cline{2-2} 
  & $\square$ simulation design (incl. pseudo-code or flow diagram)                                        \\
  & $\square$ required level of precision (incl. number of simulation   repetitions)                       \\ \hline
$\square$ & \textbf{Data-generating mechanisms}                                                            \\
  & e.g. data generation and missingness generation                                                \\ \cline{2-2} 
  & $\square$ data source (incl. model-based or design-based, sampling   variance)                         \\
  & $\square$ data characteristics (incl. multivariate relations and data   structures such as clustering) \\
  & $\square$ missingness mechanisms (incl. type or functional form of the   missing data model)           \\
  & $\square$ missingness patterns (incl. missingness proportion)                                          \\ \hline
$\square$ & \textbf{Estimands}                                                                             \\
  & e.g. complete data target                                                                      \\ \hline
$\square$ & \textbf{Methods}                                                                               \\
  & e.g. missing data methods and analytic method                                                  \\ \cline{2-2} 
  & $\square$ imputation methods (incl. parameters such as the number of   imputations)                    \\
  & $\square$ estimation method (incl. reference methods such as complete   case analysis)                 \\ \hline
$\square$ & \textbf{Performance measures}                                                                  \\
  & e.g. evaluation of estimates and imputed values                                                \\ \cline{2-2} 
  & $\square$ statistical properties (incl. comparative performance, if   applicable)                      \\
  & $\square$ validity of imputations (incl. imputation model fit and   distributional characteristics)    \\ \hline
\end{tabular}
\end{table}

>>>>

One potential reason for discrepancies between imputation method performance in simulations versus applications is the use of sampling variance in the simulation design. Simulation studies on missing data methodology have the unique option to exclude conventional sampling schemes and focus on a single finite population. When a finite population is made incomplete, sampling variance must be excluded from inference evaluations and the missing data generation procedure serves as the necessary source of Monte Carlo variation \citep{vink14, raghunathan2003multiple}. \citet{vink14} propose pooling rules for multiple imputation on finite populations and demonstrate that valid inferences are then obtained by excluding the sampling variance components from the calculation of the total variance about $(Q - \bar{Q})$, where $Q$ would be the estimand and $\bar{Q} = \sum_{l=1}^m \hat{Q}/m$ is the average over the $m$ imputed estimates \citep[cf.][p. 76-77]{rubi87}. The total variance $T=B+B/m$ would then solely rely on the between imputation variance $B = \sum_{l=1}^m (\hat{Q}_l - \bar{Q})^\prime(\hat{Q}_l - \bar{Q}) / m-1$.

@@

The challenge with model-based data-generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data may not follow a known theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness would subsequently be induced, the observed parts of the data will still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good.

@@

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to mimic a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the true data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is generated randomly. Figure \ref{Fig} demonstrates this: Both univariate MAR mechanisms in Figure \ref{Fig} are induced based on the right tail, but only the mechanisms for which the \emph{variable to ampute} and the \emph{missingness covariate} have a positive correlation will yield valid MAR mechanisms. In the case of zero correlation between the data columns, iterative univariate non-MCAR missingness schemes may induce invalid missingness mechanisms that mimic MCAR. Table \ref{Tab} demonstrates this and highlights the importance of including complete case analysis as an evaluation analysis to allow for identifying such invalidly generated missingness mechanisms in simulation. GV: REF NAAR SIMULATIE In fact, this would amount to one of the special cases described by \citet{buur18} under which complete case analysis would be more efficient than imputation: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

@@

% latex table generated in R 4.2.2 by xtable 1.8-4 package
% Thu Nov 17 17:52:47 2022
\begin{table}[t!]
\caption{Inferences obtained by complete case analysis (CCA) and multiple imputation over 1000 simulations. Displayed are the bias of the mean, coverage rate (cov) of the corresponding 95\% confidence interval and the confidence interval width (ciw). Imputations are generated by Bayesian linear regression imputation.}
\centering
\begin{tabular}{rlccccccc}
  \hline
 	& &\multicolumn{3}{c}{CCA} &&\multicolumn{3}{c}{imputation}	\\\cline{3-5}\cline{7-9}
 \multicolumn{2}{c}{mechanism}	& bias & cov & ciw && bias & cov & ciw \\ 
  \hline
$[\rho = .8]$ &mcar & 0.000 & 0.940 & 0.178 & & -0.000 & 0.941 & 0.160 \\ 
$[\rho = .8]$ &mar   & 0.335 & 0.000 & 0.169 & & 0.004 & 0.954 & 0.190 \\ 
$[\rho = .0]$ &mar   & 0.001 & 0.973 & 0.180 & & 0.011 & 0.939 & 0.311 \\ 
   \hline
\end{tabular}
\label{Tab}
\end{table}

A non-ignorable or MNAR mechanism--where the unobserved values themselves may also play a role in the probability to be missing--might fall outside many studies' scope, yet could yield informative insights for daily practice. Even if an imputation method is specifically developed with ignorable missingness in mind, chances are that the method will be applied to non-ignorable empirical missingness at some point in time. After all, for every MNAR mechanism there is a MAR mechanism with equal fit and one cannot definitively verify that empirical missingness is random \citep{molenberghs2008every}. Therefore, it can be argued that MNAR is the more likely mechanism for real-life missingness scenarios. It may be wise to include MNAR missingness in the simulation study just in case: A method that performs well under MAR and some cases of MNAR may be preferable over a method that only works under MAR, and not under MNAR. Such `late-phase` methodological investigations \citep[see][]{hei22} can provide an understanding of an imputation method's robustness. Alternatively, a sensitivity analysis may provide an indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid \citep[see e.g.][part 5]{molenberghs2014handbook}.

@@

At minimum, imputation method performance should be quantified using appropriate measures. It depends on the specifics of each simulation study which performance measures would be most suitable. If the goal is inference--not prediction--the uncertainty about estimates needs to be properly quantified. To capture this uncertainty, the standard errors of the estimates should be correctly calculated, which requires multiple imputation. The aim of multiple imputation is not to reproduce the data, but to allow for obtaining valid inference given that the data are incomplete. This means that, given the framework provided by \citet{rubi87}, and for any $(Q-\bar(Q)$, statistical properties such as bias, confidence intervals, and the coverage rate of the confidence intervals should always be studied. We therefore recommend evaluating the following points:

(i) The methods should preferably be unbiased. Note that the way bias is calculated should be carefully chosen and described, since this can greatly influence the interpretation of the results. A negligible absolute bias, for example, may already yield an almost infinite relative bias if the true value of the estimand is zero. In most cases, unbiased estimation may be expected under a MCAR missingness mechanism, which can be easily verified by including MCAR missingness in the scope of missingness investigations.

(ii) The intervals around estimates should have a valid coverage of the population (i.e. true) value. Coverage of a 95\% interval should in theory be $\geq 95$\%, where a coverage rate of 95\% would be most efficient \citep[p. 589]{neym34}. Under-coverage indicates that the procedure is not confidence valid and may lead to invalid inference. Under-coverage may occur when the estimation procedure is either biased, too liberal leading to intervals that are too narrow, or both. Over-coverage, though technically not confidence invalid, would indicate that efficiency could still be gained. In such cases, a narrower interval exists for which the method would still yield confidence valid result.

@@

(iv) Resemblence to the true data values may be quantified using the root mean squared error (RMSE). We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values and it may inflate the type I error rate of statistical inferences \citep[Chapter 2.6][]{buur18}. However, if a study is aimed at obtaining predictions, and inferential validity is not of interest, the RMSE of the predicted values can be compared between methods. Then, the RMSE may yield valuable information about the methods' comparative efficiency in terms of accuracy and precision.

After establishing the statistical validity of an imputation routine on the level of the estimand, it remains paramount to also study the imputation-generating process. Omitting such investigations may yield sub-optimal results. One simple example arises often in practice: generating implausible imputed values. Even though the estimation on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny. Under many circumstances, imputation methods may be realistically expected to preserve both marginal and conditional distributions with respect to the comparative truth. Imputation methods that fail to do so, should not be considered general-purpose methods. Many techniques can yield valid inferences, but techniques that sample realistic or plausible values may be preferable in practice.

@@

\begin{table}[ht!]
\caption{Checklist for reporting on imputation methodology evaluations.}
\label{table:check}
\begin{tabular}{ll}
\hline
$\square$ & \textbf{Aims}                                                                                  \\
  & e.g. simulation scope                                                                          \\ \cline{2-2} 
  & $\square$ simulation design (incl. pseudo-code or flow diagram)                                        \\
  & $\square$ required level of precision (incl. number of simulation   repetitions)                       \\ \hline
$\square$ & \textbf{Data-generating mechanisms}                                                            \\
  & e.g. data generation and missingness generation                                                \\ \cline{2-2} 
  & $\square$ data source (incl. model-based or design-based, sampling   variance)                         \\
  & $\square$ data characteristics (incl. multivariate relations and data   structures such as clustering) \\
  & $\square$ missingness mechanisms (incl. type or functional form of the   missing data model)           \\
  & $\square$ missingness patterns (incl. missingness proportion)                                          \\ \hline
$\square$ & \textbf{Estimands}                                                                             \\
  & e.g. complete data target                                                                      \\ \hline
$\square$ & \textbf{Methods}                                                                               \\
  & e.g. missing data methods, analytic method and inference pooling rules                                    \\ \cline{2-2} 
  & $\square$ imputation methods (incl. parameters such as the number of   imputations)                    \\
  & $\square$ estimation method (incl. reference methods such as complete   case analysis)                 \\
  & $\square$ methods used to construct standard errors and confidence intervals \\ 
  & $\;\;$ $-$ Such as rules by \citet{rubi87}, \citet{bar99}, \citet{rei03}, etc.    \\ \hline
$\square$ & \textbf{Performance measures}                                                           \\      
  & e.g. evaluation of estimates and imputed values                                                \\ \cline{2-2} 
  & $\square$ statistical properties (incl. comparative performance, if   applicable)                      \\
  & $\square$ validity of imputations (incl. imputation model fit and distributional characteristics)    \\ 
  & $\;\;$ $-$ Does the imputation model fit well to the observed values? (e.g. posterior predictive check)   \\ 
  & $\;\;$ $-$ Are there unrealistic values in the data? (e.g. negative where positive expected)   \\ 
  & $\;\;$ $-$ Are there implausible combinations in the data? (e.g. pregnant grandfathers)   \\ \hline

\end{tabular}
\end{table}
